Confirmation of \srcfull{} as a long period radio transient

### 2025-05-21 paper/main.tex
Updating paper due to co-author comments.

% Abstract of the paper
\begin{abstract}
We report the discovery of several new pulses from the source \srcfull{} (\src{}), confirming it as a long period transient (LPT) with a period of ${\sim}1.16\,$hours. The pulses are significantly scattered, consistent with Galactic electron density models. Two of the new pulses also had measurable polarisation, but unlike the originally detected pulse, the polarisation angle does not behave as expected from the rotating vector model. We interpret historical non-detections of \src{} as an intrinsic intermittency that occurs on month-long timescales, and discuss possible causes. We conjecture that, like other LPTs with periods $\gtrsim 1\,$hour, \src{} is a white-dwarf binary pulsar, but note that its period is marginally shorter than the minimum known polar orbital period.
\end{abstract}

@@

\srcfull{} (hereafter, \src{}) is a transient radio source discovered in the Variables and Slow Transients (VAST) survey \citep[][hereafter \citetalias{2024MNRAS.535..909D}]{2024MNRAS.535..909D}.
A single, highly polarized pulse was observed, lasting approximately 2 minutes, exhibiting a steep spectrum ($\alpha = -3.1$, for flux density $S_\nu\propto\nu^\alpha$).
Extensive follow up in new and archival data at multiple wavelengths turned up no other pulses.
Despite this, the authors concluded that \src{} was most likely a long period transient (LPT), a newly emerging class of radio source with periods ranging from minutes to hours \citep[see e.g.][]{2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C}.

Although optical follow-up has confirmed that at least some LPTs are white dwarfs with main sequence companions \citep{deRuiter2025,2024ApJ...976L..21H,2025A&A...695L...8R}, it remains unclear if this model can account for all known LPTs \citep[e.g.][]{2022ApJ...940...72R,Lee2025}.
Unfortunately, optical follow-up is challenging for LPTs and LPT candidates obscured by Galactic foregrounds.
This is almost certainly the case for \src{}, which is situated very close to the Galactic plane ($b = -0\overset{\circ}{.}12$), and has a dispersion measure (DM) derived distance of ${\sim}4.7\,$kpc \citepalias{2024MNRAS.535..909D}; unsurprisingly, archival searches at \src{}'s location did not turn up any source down to ${\sim}23$ AB mag.
Without more detections in the radio or at other wavelengths, it seemed unlikely that the nature of \src{} could ever be resolved.

In this paper, we present new radio observations (spanning frequencies from $154\,$MHz to $1.7\,$GHz) in which \src{} was re-detected, confirming that it is indeed an LPT with a period of ${\sim}1.16\,$hours.
In \S\ref{sec:observations}, we describe the observations themselves, some of which were taken as part of independent surveys, and some of which were taken as dedicated follow-up of \src{} once it became apparent that it was still active.
In \S\ref{sec:analysis} we present a timing analysis of the new detections, as well as revised estimates of the DM and the scattering timescale.
We also include new polarization detections in which the polarisation angle (PA) behavior differs significantly from the originally detected pulse.
In \S\ref{sec:discussion} we revisit the question of the nature of \src{}, arguing in favor of a white dwarf interpretation.
The main conclusions are summarized in \S\ref{sec:summary}.

@@

\src{} was detected blindly with the Murchison Widefield Array \citep[MWA;][]{Tingay2013} in observations taken between June--September 2024 under the ``Galactic Plane Monitor'', project code G0080. The prime objective of the campaign was to search for long-period transients by scanning the Galactic plane over $285^\circ < l < 65^\circ$, $|b|<\pm15^\circ$ on a bi-weekly cadence at 185--215\,MHz \citep[see Methods of][]{2023Natur.619..487H}; a full description will be released by Hurley-Walker et al., in prep. We also observed over June--September 2022 but made no detections in these data, as reported in \citetalias{2024MNRAS.535..909D}.
Subsequent to its redetection, we commenced a Director's Discretionary Time (DDT) campaign under project code D0042, to obtain further observations, adjusting the frequency from 200.32\,MHz to 184.96\,MHz for some of the observations to attempt to improve measurements of its DM.

We imaged both the 2024 GPM data and the follow-up DDT data at the location of \src{} using \textsc{WSClean}, masking \src{} itself, forming a deep model of the sky for each observation. After subtracting this from the visibilities, we phase-rotated to \src{}, and averaged the baseline data to produce a dynamic spectrum.
% SM to write a 'joining sentence' to explain that the timing analysis was a bit more complicated because of the low frequencies and refer the reader to the relevant section -- SM (memo to self): Not that complicated because DM + tau_sc can be ignored if making ToA predictions solely based on pulses at the target frequency.

% SM: commenting the following out because it's a repeat of above...?
%The GPM also conducted observations of this region in June--September 2022, but \src{} was not detected, either by a blind search, nor by a re-examination of the data following the methods above.

% NWH to add what went into the followup observations: D0042 (both 185 and 200 MHz). SM

@@

We also re-imaged the original 2023-05-28 L-band MeerKAT observation described in Paper~1, which was taken under project code DDT-20230525-DD-01 under capture block 1685306788. We performed a similar processing of the data, and identified four pulses with TOAs consistent with an ephemeris derived from the other data. The dynamic spectrum of one of the brightest pulses is shown in \Fig~\ref{fig:scattering_DM}.

@@

%[Reference for scattering kernels:] \citep{10.1093/mnras/stw1724}.

@@

We therefore identified the times of arrival (ToAs) with the $\mu$ parameter in the fits given in \Eqn~\ref{eqn:emg}, which represents the location of the assumed Gaussian pulse before scattering.
Because the scattering time scale could not be reliably measured for all individual pulses, we obtained ToAs by assuming a fixed scattering time of $\tau_{\rm sc,1 GHz} = 70$\,ms.

@@

@@

\subsection{Polarization} \label{sec:polarization}

@@

We conclude, therefore, that \src{} is intermittent with active periods lasting on the order of months, and that the most recent activity period fortuitously started shortly after the start of the 2024 GPM observation run.
The entire set of observations presented in this work, spanning from 2023 to 2024, may represent two or three different activity windows.
This is reminiscent of the behaviour of GLEAM-X J162759.5$-$523504.3, which was ``on'' in January 2018, ``off'' in Feburary, and ``on'' again in March \citep{2022Natur.601..526H}.
The reappearance of \src{} suggests the possibility that other known intermittent LPTs (like GLEAM-X J162759.5$-$523504.3 and GCRT J1745$-$3009) may eventually do likewise.

@@

\src{}'s period and duty cycle are not dissimilar to ILT J1101+5521 and GLEAM-X J0704$-$37, which are confirmed polars (\Fig~\ref{fig:lpt_comparison}), suggesting the possibility that \src{}, too, is a WD-MD polar system.
However, its period of ${\sim}1.16\,$hours would make it shorter than the minimum known orbital period of $1.3\,$hours for confirmed polars \citep{schwope2025polarcatcatalogpolarslowaccretion}.
This makes it an interesting testbed for the suggestion advanced by \citet{2025A&A...695L...8R} that ``short LPTs'' ($P \lesssim 78\,$mins) and ``long LPTs'' ($P \gtrsim 78\,$mins) represent different classes of systems (AM CVns and CVs, respectively).
\begin{figure*}{th}
    \centering
    \includegraphics[width=0.98\linewidth]{lpt_comparison.pdf}
    \caption{Duty cycle vs radio pulse period for various long period radio emitters \citep{deRuiter2025,2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C,2005Natur.434...50H,deRuiter2025,Lee2025,2024arXiv241116606W}. The duty cycles are derived from the reported pulse widths for LPTs, which are compared to the range of the fitted, unscattered pulse widths of \src{} at 10\% of the peak. AR\,Sco \citep{2016Natur.537..374M,2018A&A...611A..66S} and other sources without well-defined duty cycles or periods \citep[e.g.][and references therein]{2021ApJ...920...45W} are not included on this plot.}
    \label{fig:lpt_comparison}
\end{figure*}

@@

In this view, the intermittency of these systems have several possible explanations.
The limited orbital phase range in which J1912$-$4410's radio pulses were observed suggests that the geometric configuration is important in the observability of radio pulsations; ``missing'' pulses may simply occur at unfavourable orbital phases.
This is readily testable with long-term monitoring---this kind of intermittency will have the same periodicity as the orbit itself.

@@

We have shown that \src{} is in fact an LPT, as conjectured in \citetalias{2024MNRAS.535..909D}).
Its period of ${\sim}1.16\,$hours is slightly shorter than the minimum known orbital period of optically confirmed polars, marking it as an interesting case study in the context of \citet{2025A&A...695L...8R}'s suggestion that `short'-LPTs and `long'-LPTs may correspond to different classes of cataclysmic variables.

@@

It can be shown that pulses scatter-broadened by a thin screen will appear to the observer as the original pulse convolved with a one-sided exponential kernel with an associated timescale $\tau$ [CITATION].
For intrinsically Gaussian pulses of scale $\sigma$, the result of this convolution is well described by the \textit{exponentially modified Gaussian} (EMG), as given in \Eqn~\ref{eqn:emg}.

In the regime $\tau \ll \sigma$, the effect of scattering is negligible and the scattered pulse resembles the original pulse.
ToAs can be obtained in the usual way, i.e., using a matched filter ``template'' constructed from the average pulse profile.

@@

\begin{figure*}{th}
    \centering
    \includegraphics[width=0.98\linewidth]{scattering_DM.png}
    \caption{Analysis of DM measurement errors if scattering is not properly taken into account. Panel (a) shows the errors measured for each defined type of ToA, normalised to the scale of the original pulse, $\sigma$. In panel (b), the predicted error in DM is shown if the ToA error (compared to a high-frequency ToA measurement) is erroneously attributed to dispersion. Panel (c) shows the rate of change of the ToA error with frequency, and panel (d) shows the error in DM measured if this rate of change (the ``slope'' of the pulse across the observed dynamic spectrum) is erroneously attributed to dispersion. In panels (c) and (d), the derived DM errors assume a pulse scale of $\sigma = 25\,$s and $\nu_s = 230\,$MHz, as estimated for \src{}. In all panels, the blue curves indicate \emph{negative} ToA errors (i.e. the measured ToAs are earlier than the pulse's unscattered mean, $\mu$), and the red curves indicate positive quantities.}
    \label{fig:scattering_DM}
\end{figure*}

>>>>

% Abstract of the paper
\begin{abstract}
We report the discovery of several new pulses from the source \srcfull{} (\src{}), confirming it as a long period transient (LPT) with a period of ${\sim}1.16\,$hours. The pulses are significantly scattered, consistent with Galactic electron density models. Two of the new pulses also had measurable polarisation, but unlike the originally detected pulse, the polarisation angle does not behave as expected from the rotating vector model. We interpret historical non-detections of \src{} as an intrinsic intermittency that occurs on month-long timescales, and discuss possible causes. We conjecture that, like other LPTs with periods $\gtrsim 1\,$hour, \src{} is a white-dwarf binary pulsar, but note that its period is marginally shorter than the minimum known orbital period for the class of cataclysmic variables known as polars.
\end{abstract}

@@

\srcfull{} (hereafter, \src{}) is a transient radio source discovered in the Variables and Slow Transients (VAST) survey \citep[][hereafter \citetalias{2024MNRAS.535..909D}]{2024MNRAS.535..909D}.
A single, highly polarised pulse was observed, lasting approximately 2 minutes, exhibiting a steep spectrum ($\alpha = -3.1$, for flux density $S_\nu\propto\nu^\alpha$).
Extensive follow up in new and archival data at multiple wavelengths turned up no other pulses.
Despite this, the authors concluded that \src{} was most likely a long period transient (LPT), a newly emerging class of radio source with periods ranging from minutes to hours \citep[see e.g.][]{2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C}.

Although optical follow-up has confirmed that at least some LPTs are a type of strongly magnetic cataclysmic variable (CV) known as polars \citep{deRuiter2025,2024ApJ...976L..21H,2025A&A...695L...8R}, it remains unclear if this model can account for all known LPTs \citep[e.g.][]{2022ApJ...940...72R,Lee2025}.
Optical follow-up is challenging for \src{}, which is situated very close to the Galactic plane ($b = -0\overset{\circ}{.}12$), at a distance of ${\sim}4.7\,$kpc \citepalias{2024MNRAS.535..909D}.
Unsurprisingly, archival searches at \src{}'s location did not turn up any source down to ${\sim}23$ AB mag.
Without more detections in the radio or at other wavelengths, it seemed unlikely that the nature of \src{} could ever be resolved.

In this paper, we present new radio observations (spanning frequencies from $154\,$MHz to $1.7\,$GHz) in which \src{} was re-detected, confirming that it is indeed an LPT with a period of ${\sim}1.16\,$hours.
In \S\ref{sec:observations}, we describe the observations themselves, some of which were taken as part of independent surveys, and some of which were taken as dedicated follow-up of \src{} once it became apparent that it was still active.
In \S\ref{sec:analysis} we present a timing analysis of the new detections, as well as revised estimates of the dispersion measure (DM) and the scattering timescale.
We also include new polarisation detections in which the polarisation angle (PA) behaviour differs significantly from the originally detected pulse.
In \S\ref{sec:discussion} we revisit the question of the nature of \src{}, arguing in favour of a white dwarf interpretation.
The main conclusions are summarised in \S\ref{sec:summary}.

@@

\src{} was detected blindly with the Murchison Widefield Array \citep[MWA;][]{Tingay2013} in observations taken between June--September 2024 under the `Galactic Plane Monitor', project code G0080. The prime objective of the campaign was to search for long-period transients by scanning the Galactic plane over $285^\circ < l < 65^\circ$, $|b|<\pm15^\circ$ on a bi-weekly cadence at 185--215\,MHz \citep[see Methods of][]{2023Natur.619..487H}; a full description will be released by Hurley-Walker et al., in prep. We also observed over June--September 2022 but made no detections in these data, as reported in \citetalias{2024MNRAS.535..909D}.
Subsequent to its redetection, we commenced a Director's Discretionary Time (DDT) campaign under project code D0042, to obtain further observations, adjusting the frequency from 200.32\,MHz to 184.96\,MHz for some of the observations to attempt to improvehttps://github.com/CIRA-Pulsars-and-Transients-Group/vcsbeam/pull/71 measurements of its DM.
Time of arrival (ToA) predictions for the DDT observations were made using a preliminary ephemeris derived from the GPM detections, which, by virtue at being at the same or similar frequencies, absorbed the delays due to dispersion and scattering into the `PEPOCH' term.

We imaged both the 2024 GPM data and the follow-up DDT data at the location of \src{} using \citep[\textsc{WSClean}][]{2014MNRAS.444..606O}, masking \src{} itself, forming a deep model of the sky for each observation. After subtracting this from the visibilities, we phase-rotated to \src{}, and averaged the baseline data to produce a dynamic spectrum.

@@

We also re-imaged the original 2023-05-28 L-band MeerKAT observation described in \citetalias{2024MNRAS.535..909D}, which was taken under project code DDT-20230525-DD-01 under capture block 1685306788. We performed a similar processing of the data, and identified four pulses with ToAs consistent with an ephemeris derived from the other data. The folded and stacked dynamic spectra are shown in \Fig~\ref{fig:scattering_DM}.

@@

@@

We therefore identified the ToAs with the $\mu$ parameter in the fits given in \Eqn~\ref{eqn:emg}, which represents the location of the assumed Gaussian pulse before scattering.
Because the scattering time scale could not be reliably measured for all individual pulses, we obtained ToAs by assuming a fixed scattering time of $\tau_{\rm sc,1 GHz} = 70$\,ms.

@@

The variability of the pulse brightness, coupled with the fact that none of our detections were simultaneous at multiple telescopes, prohibit a direct spectral index measurement.
We note, however, that the fitted peak flux densities of the ASKAP and MeerKAT pulses are broadly consistent with the MWA flux densities if an approximate spectral index of $-2.0$ is assumed.
The in-band spectral index measurement of $-3.1$ reported for the 2023 ASKAP pulse \citepalias{2024MNRAS.535..909D}, if consistent across the span of our observations, implies a turnover or spectral curvature at intermediate frequencies \citep[perhaps comparable with GPM\,1839$-$10][]{2023Natur.619..487H}.

@@

\subsection{Polarisation} \label{sec:polarisation}

@@

We conclude, therefore, that \src{} is intermittent with active periods lasting on the order of months, and that the most recent activity period fortuitously started shortly after the start of the 2024 GPM observation run.
The entire set of observations presented in this work, spanning from 2023 to 2024, may represent two or three different activity windows.
This is reminiscent of the behaviour of GLEAM-X J162759.5$-$523504.3, which was `on' in January 2018, `off' in Feburary, and `on' again in March \citep{2022Natur.601..526H}.
The reappearance of \src{} suggests the possibility that other known intermittent LPTs (like GLEAM-X J162759.5$-$523504.3 and GCRT J1745$-$3009) may eventually do likewise.

@@

\src{}'s period and duty cycle are not dissimilar to ILT J1101+5521 and GLEAM-X J0704$-$37, which are confirmed polars (\Fig~\ref{fig:lpt_comparison}), suggesting the possibility that \src{}, too, is a WD-MD polar system.
However, its period of ${\sim}1.16\,$hours would make it shorter than the minimum known orbital period of $1.3\,$hours for confirmed polars \citep{schwope2025polarcatcatalogpolarslowaccretion}.
This makes it an interesting test bed for the suggestion advanced by \citet{2025A&A...695L...8R} that `short LPTs' ($P \lesssim 78\,$mins) and `long LPTs' ($P \gtrsim 78\,$mins) represent different classes of systems (AM Canum Venaticorum stars and CVs, respectively).
\begin{figure*}{th}
    \centering
    \includegraphics[width=0.98\linewidth]{lpt_comparison.pdf}
    \caption{Duty cycle vs radio pulse period for various long period radio emitters \citep{deRuiter2025,2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C,2005Natur.434...50H,deRuiter2025,Lee2025,2024arXiv241116606W}. The duty cycles are derived from the reported pulse widths for LPTs, which are compared to the range of the fitted, unscattered pulse widths of \src{} at 10\% of the peak. AR\,Sco \citep{2016Natur.537..374M,2018A&A...611A..66S} and other sources without well-defined duty cycles or periods \citep[e.g.][and references therein]{2021ApJ...920...45W} are not included on this plot.}
    \label{fig:lpt_comparison}
\end{figure*}

@@

In this view, the intermittency of these systems have several possible explanations.
The limited orbital phase range in which J1912$-$4410's radio pulses were observed suggests that the geometric configuration is important in the observability of radio pulsations; `missing' pulses may simply occur at unfavourable orbital phases.
This is readily testable with long-term monitoring---this kind of intermittency will have the same periodicity as the orbit itself.

@@

We have shown that \src{} is in fact an LPT, as conjectured in \citetalias{2024MNRAS.535..909D}).
Its period of ${\sim}1.16\,$hours is slightly shorter than the minimum known orbital period of optically confirmed polars, marking it as an interesting case study in the context of \citet{2025A&A...695L...8R}'s suggestion that `short'-LPTs and `long'-LPTs may correspond to different classes of CVs.

@@

It can be shown that pulses scatter-broadened by a thin screen will appear to the observer as the original pulse convolved with a one-sided exponential kernel with an associated timescale $\tau$ \citep{1972MNRAS.157...55W,1973MNRAS.163..345W}.
For intrinsically Gaussian pulses of scale $\sigma$, the result of this convolution is well described by the \textit{exponentially modified Gaussian} (EMG), as given in \Eqn~\ref{eqn:emg}.

In the regime $\tau \ll \sigma$, the effect of scattering is negligible and the scattered pulse resembles the original pulse.
ToAs can be obtained in the usual way, i.e., using a matched filter `template' constructed from the average pulse profile.

@@

\begin{figure*}{th}
    \centering
    \includegraphics[width=0.98\linewidth]{scattering_DM.png}
    \caption{Analysis of DM measurement errors if scattering is not properly taken into account. Panel (a) shows the errors measured for each defined type of ToA, normalised to the scale of the original pulse, $\sigma$. In panel (b), the predicted error in DM is shown if the ToA error (compared to a high-frequency ToA measurement) is erroneously attributed to dispersion. Panel (c) shows the rate of change of the ToA error with frequency, and panel (d) shows the error in DM measured if this rate of change (the `slope' of the pulse across the observed dynamic spectrum) is erroneously attributed to dispersion. In panels (c) and (d), the derived DM errors assume a pulse scale of $\sigma = 25\,$s and $\nu_s = 230\,$MHz, as estimated for \src{}. In all panels, the blue curves indicate \emph{negative} ToA errors (i.e. the measured ToAs are earlier than the pulse's unscattered mean, $\mu$), and the red curves indicate positive quantities.}
    \label{fig:scattering_DM}
\end{figure*}
